---
title: "OSTA Background - section 8 (Python interoperability)"
format: html
---

# 8  Python interoperability
Reference URL: https://bioconductor.org/books/release/OSTA/pages/bkg-python-interoperability.html

## Introduction
Methods discussed in this book are usually available as either R or Python packages. Ideally, users should be able to leverage the full range of available tools for their analyses. Methods should be selected based on scientific merit (ideally demonstrated by neutral benchmarks), and independent of having been implemented in a given programming language (R or Python) or framework (e.g. Bioconductor or `Seurat`).

For single-cell and spatial omics data analysis, being able to leverage different ecosystems is especially powerful. On the one hand, Python offers superb infrastructure for image analysis and machine learning-based approaches. On the other hand, the R programming language has been historically dedicated to statistical computing; as a result, many modern R methods for spatial omics data build on a solid foundation of tools for spatial statistics and statistical modeling in general.
R’s application to statistical analysis of spatial data dates back decades, primarily in epidemiological and geospatial research. As a result, various tools for spatial analyses have been established. For example, `sp `provides a “coherent set of classes and methods for […] points, lines, polygons, and grids”; `spatstat` and, more recently, `sf` provide tools for spatial point pattern and vector data, respectively.

Different data structures, although standardized within a given framework, make switching between languages and tools somewhat cumbersome. In the realm of single-cell and spatial omics, all Bioconductor tools are built around SummarizedExperiment-derived classes, while Seurat (Hao et al. 2023), Giotto (Chen et al. 2025), and VoltRon each rely on their own object definitions. In Python, Scanpy (Wolf, Angerer, and Theis 2018) and Squidpy (Palla et al. 2022) use AnnData. Attempts to alleviate the problem are being made – e.g. `zellkonverter`, `anndataR` (Deconinck et al. 2026), and functions from `Seurat` allow for conversion between Python’s AnnData and R/Bioconductor’s `SingleCellExperiment` or `SpatialExperiment`.

On a higher level, tools that enable interoperability between programming languages have become available. For example, reticulate provides an R interface to Python, including support to translate between objects from both languages; basilisk facilitates Python environment management within the Bioconductor ecosystem, and can also be interfaced with `reticulate`; and Quarto can generate dynamic reports from code in different languages.

In this chapter, we demonstrate examples showing how to set up Python environments and interact with anndata objects in R, using either a virtual or Conda environment, together with `reticulate` or `basilisk` and `anndataR`.

```{r}
#| label: dependencies
library(anndataR)
library(basilisk)
library(jsonlite)
library(VisiumIO)
library(OSTA.data)
library(reticulate)
library(SpatialExperiment)
```

## Python environment and `reticulate`

The chunk below sets up a `reticulate` environment on-the-fly for use with this Quarto document. The environment is created in the current working directory, and will be named `OSTA`. The Python version 3.8.18 suggested by the tutorial had incompatibilities with Positron tools (2026-09). The package versions listed in the `req` vector are those that were used in the tutorial.

Set to `eval: false` after the first run.

**Note:** this was troublesome to set up in a Fedora Linux WSL2 environment, particularly with compiling Python. I updated the pinned packages to be compatible with Python 3.14.0.

```{r}
#| label: original_create_reticulate_env
#| eval: false
req <- c('access==1.1.9', 'affine==2.4.0', 'anndata==0.9.2', 'anndata2ri==1.3.1', 'attrs==23.1.0', 'backports.zoneinfo==0.2.1', 'beautifulsoup4==4.12.2', 'certifi==2023.11.17', 'cffi==1.16.0', 'charset-normalizer==3.3.2', 'click==8.1.7', 'click-plugins==1.1.1', 'cligj==0.7.2', 'commot==0.0.3', 'contourpy==1.1.1', 'cycler==0.12.1', 'decorator==4.4.2', 'deprecation==2.1.0', 'esda==2.4.3', 'fiona==1.9.5', 'fonttools==4.45.1', 'gensim==4.3.2', 'geopandas==0.13.2', 'get-annotations==0.1.2', 'giddy==2.3.4', 'h5py==3.10.0', 'idna==3.4', 'igraph==0.10.8', 'importlib-metadata==6.8.0', 'importlib-resources==6.1.1', 'inequality==1.0.0', 'Jinja2==3.1.2', 'joblib==1.3.2', 'karateclub==1.3.3', 'kiwisolver==1.4.5', 'leidenalg==0.10.1', 'Levenshtein==0.23.0', 'libpysal==4.7.0', 'llvmlite==0.41.1', 'mapclassify==2.6.0', 'MarkupSafe==2.1.3', 'matplotlib==3.7.4', 'mgwr==2.1.2', 'momepy==0.6.0', 'mpmath==1.3.0', 'natsort==8.4.0', 'networkx==2.6.3', 'numba==0.58.1', 'numpy==1.22.4', 'packaging==23.2', 'pandas==1.3.5', 'patsy==0.5.3', 'Pillow==10.1.0', 'platformdirs==4.0.0', 'plotly==5.18.0', 'pointpats==2.4.0', 'POT==0.9.1', 'PuLP==2.7.0', 'pycparser==2.21', 'pygeos==0.14', 'PyGSP==0.5.1', 'pynndescent==0.5.11', 'pyparsing==3.1.1', 'pyproj==3.5.0', 'pysal==23.7', 'python-dateutil==2.8.2', 'python-igraph==0.10.8', 'python-Levenshtein==0.23.0', 'python-louvain==0.16', 'pytz==2023.3.post1', 'quantecon==0.7.1', 'rapidfuzz==3.5.2', 'rasterio==1.3.9', 'rasterstats==0.19.0', 'requests==2.31.0', 'rpy2==3.5.14', 'Rtree==1.1.0', 'scanpy==1.9.6', 'scikit-learn==1.3.2', 'scipy==1.10.1', 'seaborn==0.12.2', 'segregation==2.5', 'session-info==1.0.0', 'setuptools==68.2.2', 'shapely==2.0.2', 'simplejson==3.19.2', 'six==1.16.0', 'smart-open==6.4.0', 'snuggs==1.4.7', 'soupsieve==2.5', 'spaghetti==1.7.4', 'spglm==1.0.8', 'spint==1.0.7', 'splot==1.1.5.post1', 'spopt==0.5.0', 'spreg==1.4', 'spvcm==0.3.0', 'statsmodels==0.14.0', 'stdlib-list==0.10.0', 'sympy==1.12', 'tenacity==8.2.3', 'texttable==1.7.0', 'threadpoolctl==3.2.0', 'tobler==0.9.0', 'tqdm==4.66.1', 'tzlocal==5.2', 'umap-learn==0.5.5', 'urllib3==2.1.0', 'zipp==3.17.0')
install_python(version=ver <- "3.8.18")
virtualenv_create(envname=env <- "OSTA", python=ver, packages=req) # this is created in ~/.virtualenvs/OSTA by default
```
### Updated Python environment creation

There is a Fedora packaging quirk — Fedora splits `pip` into a separate `python3-pip` package, and reticulate's `virtualenv_starter()` validation checks for a usable `pip`, not just `venv`, before considering an interpreter "suitable." The suggested fix: `sudo dnf install python3-pip`.

Fedora also ships Python's C headers (`Python.h`) in a separate `python3.14-devel` package. Without it, `pip install` fails to compile any package with native extensions — in this list, `rpy2`, `fiona`, and (unrelatedly, see below) `gensim`. Fix: `sudo dnf install python3.14-devel`. Note that reticulate's own error message for a failed `virtualenv_create()` is unhelpfully terse (it just lists every requested package as failed); to see the real pip error, reproduce with `python3.14 -m venv` + `pip install -r <file>` directly in a shell.

`gensim==4.4.0` (the latest release on PyPI) does not build against Python 3.14's C API at all — its Cython-generated C code references CPython internals (`PyLongObject.ob_digit`, old `_PyLong_AsByteArray` signature) that were removed/changed in 3.10+. This is a genuine upstream incompatibility, independent of environment setup. `gensim` was only pulled in transitively via `karateclub` (used by `commot`, for cell-cell communication analysis), which isn't otherwise used in this document, so `commot`, `gensim`, and `karateclub` are dropped from `req_current` below.

`ipykernel` must be installed **inside** the target virtualenv, not left for Positron's bundled fallback. When a Python interpreter selected in Positron lacks its own `ipykernel`, Positron's Python extension falls back to a minimal `ipykernel` support bundle it ships internally (including its own, older `typing_extensions.py`), and prepends that path onto `sys.path`. If the venv's real `typing_extensions` is newer than the bundled one, this substitution breaks packages that rely on newer `typing_extensions` features — e.g. `import scanpy` failed here with `ImportError: cannot import name 'sentinel' from 'typing_extensions'` (via `anndata`), because Positron's bundled `typing_extensions.py` predates `typing_extensions.sentinel` and was silently shadowing the venv's own copy. Fix: add `ipykernel` to `req_current` (done below) so the venv has its own kernel support and Positron doesn't need to substitute anything.

Set to `eval: false` after the first run.

```{r}
#| label: create_reticulate_env_current
#| eval: false
# Pinned equivalents of `req`, targeting current (2026-09) releases that are
# actually installable against Python 3.14.0, rather than the tutorial's
# original Python 3.8 pins. Versions below were captured from a successful
# `pip install --dry-run` resolution against a Python 3.14.0 venv, so this
# combination of pins is known to resolve without conflicts.
# `backports.zoneinfo` is dropped: it only backports the stdlib `zoneinfo`
# module for Python <3.9 and is unbuildable/unneeded on modern CPython.
# `pygeos` is dropped: its functionality was merged into `shapely>=2.0`
# (already listed here) and the standalone package no longer builds against
# current Python/setuptools.
# `numpy` is pinned to 2.3.5 rather than the latest 2.5.3: reticulate 1.46.0
# fails to detect numpy 2.5.3 (`py_config()` reports `numpy: [NOT FOUND]`
# and `py_numpy_available()` returns FALSE) even though it imports fine from
# Python directly. This silently disables numpy-array-to-R-vector
# conversion, so e.g. `py$ad$obs$array_row` comes back as an opaque,
# unconvertible `numpy.ndarray` object instead of a numeric vector, and
# `range()`/`min()` on it fails with "invalid 'type' (list) of argument".
# numpy 2.3.5 is the earliest release with prebuilt Python 3.14 (cp314)
# wheels, and reticulate correctly detects and converts arrays from it;
# verified end-to-end with `range(py$ad$obs$array_row)` succeeding directly
# (no `$tolist()` workaround needed).
req_current <- c('access==1.1.10.post3', 'affine==3.0.1', 'anndata==0.13.3.post0', 'anndata2ri==2.0.1', 'attrs==26.1.0', 'beautifulsoup4==4.15.0', 'certifi==2026.7.22', 'cffi==2.1.1', 'charset-normalizer==3.5.1', 'click==8.5.0', 'click-plugins==1.1.1.2', 'cligj==0.7.2', 'contourpy==1.3.3', 'cycler==0.12.1', 'decorator==4.4.2', 'deprecation==2.1.0', 'esda==2.10.0', 'fiona==1.10.1', 'fonttools==4.64.0', 'geopandas==1.1.4', 'get-annotations==0.1.2', 'giddy==2.3.9', 'h5py==3.16.0', 'idna==3.19', 'igraph==1.0.0', 'importlib-metadata==9.0.1', 'importlib-resources==7.1.0', 'inequality==1.1.2', 'Jinja2==3.1.6', 'joblib==1.6.0', 'kiwisolver==1.5.1', 'leidenalg==0.12.0', 'Levenshtein==0.27.4', 'libpysal==4.15.0', 'llvmlite==0.49.0', 'mapclassify==2.11.0', 'MarkupSafe==3.0.3', 'matplotlib==3.11.1', 'mgwr==2.2.1', 'momepy==1.0.0', 'mpmath==1.3.0', 'natsort==8.4.0', 'networkx==3.6.1', 'numba==0.67.0', 'numpy==2.3.5', 'packaging==26.3', 'pandas==3.0.5', 'patsy==1.0.3', 'Pillow==12.3.0', 'platformdirs==4.11.7', 'plotly==7.0.0', 'pointpats==2.6.0', 'POT==0.9.7.post1', 'PuLP==3.3.2', 'pycparser==3.0', 'PyGSP==0.6.1', 'pynndescent==0.6.0', 'pyparsing==3.3.2', 'pyproj==3.8.0', 'pysal==26.7', 'python-dateutil==2.9.0.post0', 'python-igraph==1.0.0', 'python-Levenshtein==0.27.4', 'python-louvain==0.16', 'pytz==2026.3.post1', 'quantecon==0.11.4', 'rapidfuzz==3.14.6', 'rasterio==1.5.1', 'rasterstats==0.21.0', 'requests==2.34.2', 'rpy2==3.6.7', 'Rtree==1.4.1', 'scanpy==1.12.4', 'scikit-learn==1.9.0', 'scipy==1.18.1', 'seaborn==0.13.2', 'segregation==2.5.5', 'session-info==1.0.1', 'setuptools==84.0.0', 'shapely==2.1.2', 'simplejson==4.1.2', 'six==1.17.0', 'smart-open==8.0.1', 'snuggs==1.4.7', 'soupsieve==2.9.2', 'spaghetti==1.7.6', 'spglm==1.1.0', 'spint==1.1.0', 'splot==1.1.7', 'spopt==0.7.0', 'spreg==1.9.1', 'spvcm==0.3.0', 'statsmodels==0.15.0', 'stdlib-list==0.12.0', 'sympy==1.14.0', 'tenacity==9.1.4', 'texttable==1.7.0', 'threadpoolctl==3.6.0', 'tobler==0.14.0', 'tqdm==4.70.0', 'tzlocal==5.4.4', 'umap-learn==0.5.12', 'urllib3==2.7.0', 'zipp==4.1.0', 'ipykernel', 'squidpy==1.8.3')
# # Resolve the path to an existing Python 3.14 interpreter on this machine
# rather than hardcoding a path, so reticulate/pyenv does not attempt to
# download and build a separate copy, and so this resolves portably on
# other machines with a system Python 3.14 already installed.
virtualenv_create(envname=env_current <- "OSTA_current", python=ver_current <- virtualenv_starter("3.14"), packages=req_current)

# Positron quirk workaround (see AGENTS.md item 6): Positron always injects
# PYTHONPATH entries pointing at its own bundled ipykernel support libraries
# ahead of this venv's own site-packages, which can shadow newer packages
# (e.g. an older bundled `typing_extensions.py` missing `sentinel`, breaking
# `import scanpy`/`anndata`). A `sitecustomize.py` dropped into this venv's
# site-packages re-prioritizes the venv's own packages at interpreter
# startup, before any user code runs. This lives inside the venv (not this
# repo), so it must be (re-)written any time the venv is (re-)created --
# hence doing it here, immediately after `virtualenv_create()`.
venv_python <- virtualenv_python(env_current)
site_packages <- system2(venv_python, c("-c", shQuote("import sysconfig; print(sysconfig.get_path('purelib'))")), stdout=TRUE)
sitecustomize_path <- file.path(site_packages, "sitecustomize.py")
writeLines(c(
  '"""',
  "Workaround for a Positron/reticulate interaction: Positron's Python",
  "extension always injects PYTHONPATH entries pointing at its own bundled",
  "ipykernel support libraries, placed ahead of this virtualenv's own",
  "site-packages on sys.path -- even when this venv has its own ipykernel",
  "installed. Those bundled libraries can ship an older `typing_extensions.py`",
  "missing symbols (e.g. `sentinel`) needed by newer packages such as",
  "`anndata`, silently shadowing this venv's own, newer `typing_extensions`",
  "install and breaking imports like `import scanpy`.",
  "",
  "Fix: move this venv's own site-packages directory to the front of",
  "`sys.path`, so it takes priority over Positron's bundled fallback",
  "libraries. See AGENTS.md in the OSTA_walkthroughs project for the full",
  "quirk write-up. This file is regenerated by the",
  "`create_reticulate_env_current` chunk in background_section8.qmd every",
  "time the venv is (re-)created.",
  '"""',
  "",
  "import sys",
  "import os",
  "",
  "_venv_site_packages = os.path.dirname(os.path.abspath(__file__))",
  "",
  "if _venv_site_packages in sys.path:",
  "    sys.path.remove(_venv_site_packages)",
  "sys.path.insert(0, _venv_site_packages)"
), sitecustomize_path)
```

### Use the previously created virtual environment
```{r}
#| eval: true
use_virtualenv("OSTA_current", required=TRUE)
```

## Alternative: use conda

In addition, I constructed a conda environment route (2026-09). There is a setup script in this repository `setup_osta_env.sh` for this purpose.

As of 2026-09-06 this may require additional work to achieve parity with the virtualenv route, due to extensive work necessary to make Positron operate correctly with the virtual environment after the conda script was created.

Set as `eval: false` by default

```{r}
#| label: conda_env
#| eval: false

# path to Conda binary
bin <- ".../bin/conda" 
options(reticulate.conda_binary=bin)
# activate environment
use_condaenv("<name>")
```

## Interfacing between languages

### SingleCellExperiment

After configuring Python, R commands can now be run using `reticulate`. First, the AnnData object is imported using `scanpy` (original package used by tutorial) or `squidpy`. This inovlves a download from exampledata.scverse.org

```{python}
#| label: import_anndata
# import scanpy
import squidpy
id = "V1_Mouse_Brain_Sagittal_Posterior"
# ad = scanpy.datasets.visium_sge(sample_id=id) # this was the original tutorial command but 2026-09-06 deprecation notice recommend squidpy command below
ad = squidpy.datasets.visium(sample_id=id)
ad.var_names_make_unique()
ad
```

Back to R

```{r}
range(py$ad$obs$array_row)
```
